---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
options(readr.show_col_types = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# datagouv

<!-- badges: start -->
[![R-CMD-check](https://github.com/astamm/datagouv/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/astamm/datagouv/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/astamm/datagouv/graph/badge.svg)](https://app.codecov.io/gh/astamm/datagouv)
[![pkgdown](https://github.com/astamm/datagouv/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/astamm/datagouv/actions/workflows/pkgdown.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/datagouv)](https://CRAN.R-project.org/package=datagouv)
<!-- badges: end -->

`datagouv` is an R client for the public API of
[data.gouv.fr](https://www.data.gouv.fr), the French government's open data
platform. It helps you *find* a dataset that matches your interests, *judge*
whether it is usable, *download* it, and later *re-fetch the exact same table*
reproducibly. Requests are built on top of the
[`httr2`](https://httr2.r-lib.org) package.

## Installation

```r
# From GitHub once the package is published
# remotes::install_github("astamm/datagouv")
```

## Try it in seconds

The examples below hit the live data.gouv.fr API and show real results.

Find datasets matching a topic — `dg_list_datasets()` searches titles and
descriptions server-side:

```{r}
library(datagouv)

hits <- dg_list_datasets(q = "vélo", n = 5)
hits[, c("title", "formats", "n_resources", "has_table", "has_schema")]
```

`id` is the stable identifier you use to download; `has_schema` tells you
whether the dataset declares per-column documentation (see below).

Download a dataset and inspect it. Each returned table carries a trailing `.id`
column — its stable address:

```{r}
tables <- dg_pull_dataset("6397c0ff56d3963118a18345")   # C.vélo bike stations
head(tables[[1]][, 1:6])
```

Judge whether the columns mean what you think. `dg_schema()` resolves the
dataset's declared schema and returns human-readable column documentation:

```{r}
# Find a schema-documented dataset, then look at its documented columns.
irve <- dg_pull_dataset("6a84778d27ac6d44d5fabe1f")     # IRVE charging points
dg_schema(irve[[1]]$.id[1])[, c("name", "description", "type", "example")]
```

Re-fetch the exact same table later, reproducibly, from its stored `.id` — no
matter how the catalog changes underneath you:

```{r}
table_id <- tables[[1]]$.id[1]
again <- dg_refetch(table_id)

identical(again, tables[[1]])
```

And compute summary metrics — size, columns, rows and missing-value rate — on
one table or several at once:

```{r}
summarise_datasets(datasets = list(iris = iris, mtcars = mtcars))

out <- dg_download_many(c("6397c0ff56d3963118a18345"))
out$metrics
```

## Supported formats

`dg_pull_dataset()` downloads the first tabular resource of a dataset among
CSV, CSV.GZ, XLS, XLSX, PARQUET, TSV, TXT and JSON, and returns the parsed
table(s) as a named list. A ZIP resource is unpacked and every contained file
in one of these formats is parsed — one element of the list per file. The
delimiter of CSV/TXT resources is auto-detected (comma, semicolon, tab, pipe,
...), so both standard and European-style (semicolon/comma-decimal) files are
handled without special configuration.

The discovery catalog (`dg_list_datasets()`) is restricted to the official
tabular formats data.gouv.fr itself indexes (`csv`, `csv.gz`, `xls`, `xlsx`,
`parquet`), so every listed dataset is in principle openable as a table. Direct
pulls additionally accept `tsv`, `txt` and `json` resources.

See the [vignette](https://astamm.github.io/datagouv/articles/datagouv.html) for
the full workflow and API reference.

## Disclaimer

This package is not affiliated with or endorsed by data.gouv.fr. It is an
independent project, and the authors are not responsible for the content of
the datasets it indexes. This first version is a proof of concept and may
contain bugs. Please report any issues on the
[GitHub repository](https://github.com/astamm/datagouv/issues).

This package is the result of a joint effort initiated during the French
Finist'R bootcamp which was held in Roscoff in August 2026. It served as a
complex task to benchmark R package development assisted by AI. The refinements
that came up during the bootcamp were made with the assistance of
`deepseek-v4-flash` as provided by Albert API, the French government's AI
platform. The author is grateful to the Finist'R organizers and the Albert API
team for their support and guidance.