---
title: "Finding, judging and re-fetching French open data with datagouv"
author: "Aymeric Stamm"
format: html
editor: source
vignette: >
  %\VignetteIndexEntry{Finding, judging and re-fetching French open data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

> This vignette assumes you can reach the data.gouv.fr API. Code that touches
> the live API only runs in an interactive session so that the vignette builds
> cleanly; the worked examples below show the expected output.

The examples that hit the live API are marked `#| live: true`; they run only in
an interactive session (e.g. when building this site with pkgdown), and are
skipped during `R CMD check`. In-memory examples run unconditionally:

```{r}
#| include: false
library(datagouv)
knitr::opts_hooks$set(live = function(options) {
  if (isTRUE(options$live)) options$eval <- interactive()
  options
})
```

## The problem datagouv solves

`datagouv` is a small R client for the public API of
[data.gouv.fr](https://www.data.gouv.fr), the French government's open data
platform. It was written with a specific user in mind: a student or a data
scientist who wants to *find* a dataset that matches their interests, *judge*
whether it is usable for analysis, *download* it, and later *re-fetch the exact
same table* in a reproducible way.

These four steps may sound trivial, but the platform makes each of them harder
than it should be:

1. **Finding.** Enumerating "all" published datasets means paging through
   tens of thousands of records. The catalog search is server-side and only
   matches titles and descriptions, which are free-form and often uninformative.
2. **Judging.** The tabular API provides no human-readable variable
   descriptions. The actual descriptions live in separate *schema* documents on
   [schema.data.gouv.fr](https://schema.data.gouv.fr); data.gouv.fr only attaches
   a pointer (`name` / `url`) to a resource. Without resolving that pointer, you
   cannot tell whether a column really means what you assume.
3. **Downloading.** Datasets come in many formats (CSV, Excel, ZIP, Parquet,
   JSON...), the declared format is not always accurate, and a "dataset" may
   contain several files.
4. **Re-fetching.** Catalog titles and even ordering change over time, so a
   script that reaches for a dataset "by title" is fragile. You want an address
   that always resolves to the same table.

`datagouv` addresses all four. Its functions are organised around that workflow:

| Step | Function |
|------|----------|
| Find / search the catalog | `dg_list_datasets()` |
| Judge documented columns | `dg_schema()` |
| Download tabular resources | `dg_pull_dataset()` |
| Summarise table contents | `get_summary()`, `summarise_datasets()` |
| Re-fetch a table reproducibly | `dg_refetch()` |
| Download several at once | `dg_download_many()` |

The approach mirrors what several US cities propose (e.g.
[`nycOpenData`](https://github.com/ropensci/nycOpenData) for New York), but
tailored to the data.gouv.fr API.

## Finding datasets

`dg_list_datasets()` returns a tibble with one row per dataset:

```{r}
#| live: true
library(datagouv)

datasets <- dg_list_datasets(n = 20)
head(datasets)
```

The columns are chosen to help you decide, at a glance, whether a dataset is
worth pulling:

* `id` is the stable, unique identifier used to address the dataset later.
* `n_resources` is the number of files the dataset contains.
* `formats` lists the distinct file formats found among them.
* `has_table` is `TRUE` when at least one resource can be parsed into a table
  by this package.
* `has_schema` is `TRUE` when at least one resource declares a data schema,
  i.e. when per-variable documentation is available via `dg_schema()`.

By default `dg_list_datasets()` returns the first `n` datasets of the catalog,
which is both slow and fragile when done exhaustively. Prefer a server-side
search with `q`:

```{r}
#| live: true
cycle <- dg_list_datasets(q = "vélo", n = 10)
cycle[, c("title", "n_resources", "has_table", "has_schema")]
```

The discovery catalog is **restricted to data.gouv's official tabular
formats** (`csv`, `csv.gz`, `xls`, `xlsx`, `parquet`), so every listed dataset
is in principle openable as a table — `has_table` is almost always `TRUE`. This
is a deliberate choice: JSON, TSV and TXT resources can still be parsed when
you address them directly (see `dg_pull_dataset()`), but they are not guaranteed
tabular and are left out of the catalog so that "the catalog" stays a reliable
list of tables.

### Restricting to documented datasets

Because descriptions live in schemas and only a fraction of datasets declare
one, `dg_schema()` only helps on a subset of the catalog. You can target that
subset directly:

```{r}
#| live: true
documented <- dg_list_datasets(schema_only = TRUE, n = 10)
documented[, c("title", "n_resources", "has_table", "has_schema")]
```

## Judging whether a dataset is usable

The judged usefulness of a dataset hinges on whether the columns mean what you
think they mean. That information comes from the producer's *schema* on
schema.data.gouv.fr. `dg_schema()` takes a composed table id (the value stored
in the `.id` column of a pulled table) and returns the documented fields:

```{r}
#| live: true
documented <- dg_list_datasets(schema_only = TRUE, n = 1)
table_id <- documented$id[[1]]

# Pull it, then inspect the schema of the returned table.
tables <- dg_pull_dataset(table_id)
schema <- dg_schema(tables[[1]]$.id[1])

# Human-readable titles and descriptions of every column:
head(schema)
```

The result is a tibble with one row per column and the columns `name`,
`title`, `description`, `type` and `example`, together with the schema's own
`title` and `name` attached as attributes. Where the schema provides no title
or description (some producers document only some fields), the corresponding
cell is `NA`.

If the resource carries no schema pointer, `dg_schema()` returns `NULL` with a
message explaining that no variable documentation is available.

## Downloading data

`dg_pull_dataset()` downloads the first parseable tabular resource of a dataset
and returns a **named list of tibbles**:

```{r}
#| live: true
tables <- dg_pull_dataset("6397c0ff56d3963118a18345")
names(tables)
head(tables[[1]])
```

A few things to know about pulling:

* Supported formats are `csv`, `csv.gz`, `xls`, `xlsx`, `parquet`, `tsv`, `txt`
  and `json`. A **ZIP** resource is unpacked and every contained file in one of
  these formats is parsed — one element of the returned list per file.
* The delimiter of CSV/TXT resources is auto-detected (comma, semicolon, tab,
  pipe, ...), so both standard and European-style files (semicolon /
  decimal-comma) are handled without configuration.
* Declared formats are not always accurate, and a candidate may fail to parse
  (e.g. a `json` resource that actually serves an API metadata document).
  `dg_pull_dataset()` skips non-parseable resources and falls back to the next
  tabular one instead of erroring.
* Every returned table carries a trailing `.id` column — its stable, unique
  address.

## Re-fetching the same table reproducibly

This is what makes your analysis reproducible over time. The `.id` column is
built from the platform's own stable identifiers
(`<dataset_id>::<resource_id>`, plus the file name for a file inside a ZIP).
Unlike a human-readable title, this address always resolves to the same table:

```{r}
#| live: true
tables <- dg_pull_dataset("6397c0ff56d3963118a18345")
table_id <- tables[[1]]$.id[1]
table_id

# Re-fetch the exact same table later:
again <- dg_refetch(table_id)
```

`dg_refetch()` accepts the composed id directly, so you can store it in a
script or a database and reproduce the pull without re-searching the catalog.

## Summarising datasets

`get_summary()` computes metrics for a single table:

```{r}
get_summary(iris, name = "iris")
```

The reported columns are `dataset` (a label), `size_kb` (in-memory weight),
`n_vars`, `n_numeric`, `n_non_numeric`, `n_rows` and `prop_missing` (the
proportion of missing values). The `.id` column added by `dg_pull_dataset()` is
excluded from these metrics — it is an address, not a data variable.

`summarise_datasets()` applies `get_summary()` to a collection of tables. It is
flexible about its input, accepting:

* a named list of tibbles (each element is a single table),
* a named list of such lists, as returned by `dg_pull_dataset()` /
  `dg_download_many()` (a ZIP may contribute several tables),
* a tibble from `dg_list_datasets()` (each dataset is downloaded and
  summarised),
* a character vector of identifiers (or exact titles),
* or `NULL` to download and summarise the first `n` datasets of the catalog.

```{r}
# In-memory tables — no network needed
summarise_datasets(datasets = list(iris = iris, mtcars = mtcars))
```

`dg_download_many()` bundles the two most common operations: download several
datasets and return both the raw tibbles and the metrics.

```{r}
#| live: true
out <- dg_download_many(c("6397c0ff56d3963118a18345"))
out$metrics
head(out$datasets[[1]])
```

## A complete workflow

Putting it all together, a typical session is:

```{r}
#| live: true
# 1. Find a documented dataset on a topic of interest.
hits <- dg_list_datasets(q = "recharge électrique", schema_only = TRUE, n = 5)
hits[, c("title", "has_table")]

# 2. Pull the first one and confirm its columns are documented.
tables <- dg_pull_dataset(hits$id[[1]])
tbl_id <- tables[[1]]$.id[1]
dg_schema(tbl_id)

# 3. Look at the table and store its stable id for later.
head(tables[[1]])

# 4. Re-fetch the exact same table in a later session.
again <- dg_refetch(tbl_id)
```

The table id is the key to reproducibility: save it, and `dg_refetch(tbl_id)`
will give you the same table again, no matter how the public catalog changes in
the meantime.
